---
title: "Biomass3"
author: "Sarai Hutchinson"
date: "2025-10-06"
output: html_document
---

```{r}
# ============================================================
# 0) Libraries & Setup
# ============================================================
library(tidyverse)  # dplyr, ggplot2, readr, tidyr, etc.
library(car)        # Levene's test
library(agricolae)  # HSD.test
library(gridExtra)  # grid.arrange
library(tibble)     # rownames_to_column
library(patchwork)
library(emmeans)
library(multcomp)
library(broom)
library(performance)

# Optional: guard against function masking (uncomment if needed)
# conflicted::conflict_prefer("filter", "dplyr")

# Palettes
okabe_ito <- c("#009E73", "#D55E00", "#E69F00", "#56B4E9", "#F0E442", "#0072B2", "#CC79A7", "#999999")
color_vals <- c(
  "Soil"                = okabe_ito[1],
  "Glass"               = okabe_ito[6],
  "25% SG"              = okabe_ito[3],
  "75% SG"              = okabe_ito[2],
  "100% SG"             = okabe_ito[7]
)

# ============================================================
# 1) Load & Inspect Raw Data
# ============================================================
data_path <- "C:\\Users\\sarai\\OneDrive\\Email attachments\\Documents\\UVI\\Thesis.Work\\SargMangs\\Data/Combined_Biomass.csv"
raw_data  <- readr::read_csv(data_path, show_col_types = FALSE)

cat("Data structure:\n"); str(raw_data)
cat("\nFirst few rows:\n"); print(head(raw_data))
cat("\nColumn names:\n"); print(names(raw_data))
cat("\nMissing values by column:\n"); print(colSums(is.na(raw_data)))
```


```{r}
# ============================================================
# 2) Tidy to Wide (per plant) & Harmonize Columns
# ============================================================
processed_raw <- raw_data %>%
  dplyr::filter(SAMPLE != "DEAD ROOTS") %>%
  dplyr::select(TREATMENT, SPECIES, PROP_ID, SAMPLE, `TOT_DRY_WT (g)`) %>%
  tidyr::pivot_wider(
    id_cols    = c(TREATMENT, SPECIES, PROP_ID),
    names_from = SAMPLE,
    values_from = `TOT_DRY_WT (g)`,
    names_repair = "minimal"
  ) %>%
  dplyr::rename_with(~ stringr::str_to_lower(stringr::str_replace_all(.x, " ", "_"))) %>%
  dplyr::rename(
    treatment    = treatment,
    species      = species,
    prop_id      = prop_id,
    leaves       = leaves,
    stem         = stem,
    fine_roots   = fine_roots,
    coarse_roots = coarse_roots
  )

# # After your pivot_wider + rename block
# # (1) Ensure hypocotyl column is present and correct per species
# if ("hypo" %in% names(processed_raw)) {
#   processed_raw <- processed_raw %>%
#     dplyr::rename(hypocotyl = hypo) %>%
#     dplyr::mutate(
#       hypocotyl = dplyr::case_when(
#         species == "LARA" ~ 0,                  # always 0 for LARA
#         is.na(hypocotyl)  ~ 0,                  # if missing for RHMA, set 0
#         TRUE               ~ hypocotyl
#       )
#     )
# } else {
#   processed_raw <- processed_raw %>%
#     dplyr::mutate(hypocotyl = dplyr::if_else(species == "LARA", 0, 0))  # 0 for all if absent
# }

# Map TREATMENT codes A–E to labels (keep existing labels if already set)
processed_raw <- processed_raw %>%
  dplyr::mutate(
    treatment = dplyr::case_when(
      treatment == "A" ~ "Soil",
      treatment == "B" ~ "Glass",
      treatment == "C" ~ "100% SG",
      treatment == "D" ~ "25% SG",
      treatment == "E" ~ "75% SG",
      TRUE ~ treatment
    )
  )

processed_clean <- processed_raw %>%
  mutate(
    # convert negatives to 0 or abs() for biological realism
    across(c(leaves, stem, fine_roots, coarse_roots),
           ~ if_else(.x < 0, abs(.x), .x)),
    data_note = if_else(
      leaves < 0 | stem < 0 | fine_roots < 0 | coarse_roots < 0,
      "Corrected negative weight (paper bag error)",
      NA_character_
    )
  )

biomass_data <- processed_clean
```

Creating a figure with all the biomass displayed plus significant letters
```{r}
# ============================================================
# 0) Libraries
# ============================================================
library(dplyr)
library(tidyr)
library(forcats)
library(ggplot2)
library(stringr)

#use abs() instead of 0, since even if the true value was small, it wasn’t literally zero — this preserves the magnitude distribution.
processed_clean <- processed_raw %>%
  mutate(
    # convert negatives to 0 or abs() for biological realism
    across(c(leaves, stem, fine_roots, coarse_roots),
           ~ if_else(.x < 0, abs(.x), .x)),
    data_note = if_else(
      leaves < 0 | stem < 0 | fine_roots < 0 | coarse_roots < 0,
      "Corrected negative weight (paper bag error)",
      NA_character_
    )
  )
# ============================================================
# 1) Options you can toggle
# ============================================================
include_hypocotyl_in_stem <- FALSE   # TRUE if you want RHMA hypocotyl added to "Stem"
treat_order <- c("Soil","Glass","25% SG","75% SG","100% SG")

# ============================================================
# 2) Clean + compute components from your processed_raw
#    Expect columns: treatment, species, prop_id, leaves, stem, fine_roots, coarse_roots
#    Optional: hypocotyl (for RHMA). If absent it’s treated as 0.
# ============================================================
dat_biomass <- processed_clean %>%
  mutate(
    leaves       = coalesce(leaves, 0),
    stem         = coalesce(stem, 0),
    fine_roots   = coalesce(fine_roots, 0),
    coarse_roots = coalesce(coarse_roots, 0),
    hypocotyl    = if ("hypocotyl" %in% names(.)) coalesce(hypocotyl, 0) else 0,
    Stem         = stem + if_else(include_hypocotyl_in_stem & species == "RHMA", hypocotyl, 0),
    Leaf         = leaves,
    Root         = fine_roots + coarse_roots,
    treatment    = factor(treatment, levels = treat_order)
  ) %>%
  dplyr::select(species, treatment, Leaf, Stem, Root)

# ============================================================
# 3) Summarize: mean ± SE by species × treatment × component
#    (Mirror Root below zero for plotting)
# ============================================================
# --- Summarize means & SEs per component (LONG) ---
# ---- 1) Summaries ----
sum_long <- dat_biomass %>%
  pivot_longer(c(Leaf, Stem, Root), names_to = "Component", values_to = "value") %>%
  group_by(species, treatment, Component) %>%
  summarise(
    n    = sum(!is.na(value)),
    mean = mean(value, na.rm = TRUE),
    se   = ifelse(n > 1, sd(value, na.rm = TRUE)/sqrt(n), NA_real_),
    .groups = "drop"
  )

sum_wide <- sum_long %>%
  pivot_wider(names_from = Component, values_from = c(mean, se), names_sep = "_")
# cols: mean_Leaf, se_Leaf, mean_Stem, se_Stem, mean_Root, se_Root

# ---- 2) Helper to build one species plot ----
build_species_plot <- function(sp) {
  # Filter species
  bars_sp <- sum_long %>% filter(species == sp)

  # Positive stack: Stem (bottom) then Leaf (top)
  pos_bars <- bars_sp %>%
    filter(Component %in% c("Leaf", "Stem")) %>%
    mutate(
      Component = factor(Component, levels = c("Leaf","Stem")),  # order = bottom->top
      y = mean
    )

  # Negative: Root mirrored
  neg_bars <- bars_sp %>%
    filter(Component == "Root") %>%
    mutate(y = -mean)

  # Error bars at true tops
  eb_sp <- sum_wide %>% filter(species == sp) %>%
    transmute(
      species, treatment,
      # tops
      top_Stem =  mean_Stem,
      top_Leaf =  mean_Stem + mean_Leaf,
      top_Root = -mean_Root,
      # ses
      se_Stem  = se_Stem,
      se_Leaf  = se_Leaf,
      se_Root  = se_Root
    ) %>%
    pivot_longer(
      cols = -c(species, treatment),
      names_to = c(".value","Component"),
      names_pattern = "(top|se)_(Leaf|Stem|Root)"
    ) %>%
    mutate(ymin = top - se,
           ymax = top + se)

  # Axis extent from real tops (incl. SE)
  axis_extent <- max(abs(c(eb_sp$ymin, eb_sp$ymax)), na.rm = TRUE)

  # Plot
  ggplot() +
    # Positive stack: Stem (bottom) then Leaf (top)
    geom_col(data = pos_bars, aes(treatment, y, fill = Component),
             width = 0.7, color = "black", linewidth = 0.25, position = "stack") +
    # Negative mirrored Root
    geom_col(data = neg_bars, aes(treatment, y, fill = Component),
             width = 0.7, color = "black", linewidth = 0.25) +
    # Error bars at tops
    geom_errorbar(data = eb_sp,
                  aes(treatment, ymin = ymin, ymax = ymax, group = Component),
                  width = 0.18, linewidth = 0.25, na.rm = TRUE) +
    geom_hline(yintercept = 0, linewidth = 0.3) +
    scale_fill_manual(values = c(Leaf = "white", Stem = "grey60", Root = "grey10"),
                      breaks = c("Leaf","Stem","Root"), name = NULL) +
    scale_y_continuous(
      limits = c(-1.10*axis_extent, 1.10*axis_extent),
      breaks = scales::pretty_breaks(5),
      labels = function(x) abs(x)  # show positives on both sides
    ) +
    labs(x = "Treatment", y = "Biomass (g)", title = sp) +
    theme_classic(base_size = 12) +
    theme(
      legend.position = "right",
      legend.background = element_rect(color = "black", fill = "white")
    )
}

# ---- 3) Build the two panels ----
p_LARA <- build_species_plot("LARA") + theme(legend.position = "none")
p_LARA
p_RHMA <- build_species_plot("RHMA")  # keep legend here
p_RHMA

# If you use patchwork:
# library(patchwork)
# p_LARA + p_RHMA + plot_layout(widths = c(1,1))

```



I got feedback from Kristin to remove hypocotyl from the above ground biomass to remove the site/ propagule effect.
```{r}
library(forcats)
biomass <- biomass_data %>%
  # compute components and totals
  dplyr::mutate(
    # Above-ground biomass = leaves + stem (+ hypocotyl for RHMA)
    # AGB        = pmax(rowSums(dplyr::across(c(leaves, stem, hypocotyl)), na.rm = TRUE), 0), #run this code if you want to add hypocotyl back to the above-ground biomass
    AGB        = pmax(rowSums(dplyr::across(c(leaves, stem)), na.rm = TRUE), 0),
    FineRoot   = pmax(fine_roots   %||% 0, 0),
    CoarseRoot = pmax(coarse_roots %||% 0, 0),
    BGB_total  = FineRoot + CoarseRoot
  ) %>%
  # harmonize factor names/levels used in analysis
  dplyr::mutate(
    Species        = forcats::fct_relevel(toupper(species), "LARA","RHMA"),
    Treatment_Name = forcats::fct_relevel(treatment, "Soil","Glass","25% SG","75% SG","100% SG")
  ) %>%
   # transformed responses for modeling (handles zeros)
  # transformed responses + log-ratios
  dplyr::mutate(
    AGB_lp    = log1p(AGB),      # log(1 + AGB)
    Fine_lp   = log1p(FineRoot),
    Coarse_lp = log1p(CoarseRoot),
    BGB_lp    = log1p(BGB_total),
    # equals log((1+BGB)/(1+AGB)); don’t report this as “raw R:S”
    logRS     = BGB_lp-AGB_lp,   # log root:shoot
    logFC     = Fine_lp-Coarse_lp
  )


# Convenience split
biomass_R <- filter(biomass, Species == "RHMA")
biomass_L <- filter(biomass, Species == "LARA")

# Helper to compute EMMeans letters (Holm-adjusted) and back-transform where needed
letters_tbl <- function(fit, resp = c("AGB_lp","Fine_lp","Coarse_lp","logRS","logFC"), transform = c("exp1","identity")){
  resp <- match.arg(resp)
  transform <- match.arg(transform)
  em <- emmeans(fit, ~ Treatment_Name)
  # compact letter display with Holm
  cl <- multcomp::cld(em, Letters = letters, adjust = "holm")
  out <- as.data.frame(cl) %>%
    rename(letters = .group) %>%
    mutate(letters = trimws(letters))
  # back-transform if requested
  if(transform == "exp1"){
    if(resp %in% c("AGB_lp","Fine_lp","Coarse_lp")){
      out <- out %>% mutate(mean_bt = exp(emmean) - 1,
                            lower_bt = exp(lower.CL) - 1,
                            upper_bt = exp(upper.CL) - 1)
    } else if(resp %in% c("logRS","logFC")){
      # For log-ratios, exponentiate to ratio scale
      out <- out %>% mutate(mean_bt = exp(emmean),
                            lower_bt = exp(lower.CL),
                            upper_bt = exp(upper.CL))
    }
  }
  out
}

```

Means by Component (AGB, Fine, Coarse)

We analyze AGB, fine-root biomass, and coarse-root biomass separately per species with fixed effects models. Responses are log1p-transformed for normality and to accommodate zeros; we back-transform for presentation. Pairwise tests use Holm adjustment and letters for compact display.

RHMA
```{r}
fit_agb_R    <- lm(AGB_lp   ~ Treatment_Name, data = biomass_R)
fit_fine_R   <- lm(Fine_lp  ~ Treatment_Name, data = biomass_R)
fit_coarse_R <- lm(Coarse_lp~ Treatment_Name, data = biomass_R)

anova(fit_agb_R); anova(fit_fine_R); anova(fit_coarse_R)

letters_agb_R    <- letters_tbl(fit_agb_R, resp="AGB_lp", transform="exp1")
letters_fine_R   <- letters_tbl(fit_fine_R, resp="Fine_lp", transform="exp1")
letters_coarse_R <- letters_tbl(fit_coarse_R, resp="Coarse_lp", transform="exp1")

```
# Analysis of Variance Table
#
# Response: AGB_lp
#                Df Sum Sq Mean Sq F value
# Treatment_Name  4 3.5937 0.89841  8.4986
# Residuals      45 4.7571 0.10571
#                   Pr(>F)
# Treatment_Name 3.389e-05 ***
# Residuals
# ---
# Signif. codes:
# 0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# Analysis of Variance Table
#
# Response: Fine_lp
#                Df Sum Sq  Mean Sq F value
# Treatment_Name  4 1.1744 0.293589  9.0602
# Residuals      45 1.4582 0.032404
#                   Pr(>F)
# Treatment_Name 1.863e-05 ***
# Residuals
# ---
# Signif. codes:
# 0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# Analysis of Variance Table
#
# Response: Coarse_lp
#                Df Sum Sq Mean Sq F value
# Treatment_Name  4 2.4857 0.62143  8.0491
# Residuals      45 3.4742 0.07720
#                  Pr(>F)
# Treatment_Name 5.53e-05 ***
# Residuals
# ---
# Signif. codes:
# 0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

```{r}
plot_box_letters <- function(df, y, ylab){
  ggplot(df, aes(Treatment_Name, .data[[y]], fill = Treatment_Name)) +
    geom_jitter(aes(color = Treatment_Name), width = 0.15, alpha = 0.6, size = 2) +
    geom_boxplot(alpha = 0.3, outlier.shape = NA, width = 0.6) +
    stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "black") +
    scale_fill_manual(values = color_vals) + scale_color_manual(values = color_vals) +
    labs(x = NULL, y = ylab) +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 15, hjust = 1))
}

p1 <- plot_box_letters(biomass_R, "AGB",       "Above-ground biomass (g)")
p2 <- plot_box_letters(biomass_R, "FineRoot",  "Fine-root biomass (g)")
p3 <- plot_box_letters(biomass_R, "CoarseRoot","Coarse-root biomass (g)")

p1 / p2 / p3

```

```{r}
letters_agb_R %>% dplyr::select(Treatment_Name, mean_bt, lower_bt, upper_bt, letters) %>% arrange(Treatment_Name)
letters_fine_R %>% dplyr::select(Treatment_Name, mean_bt, lower_bt, upper_bt, letters) %>% arrange(Treatment_Name)
letters_coarse_R %>% dplyr::select(Treatment_Name, mean_bt, lower_bt, upper_bt, letters) %>% arrange(Treatment_Name)

```
Treatment_Name
<fctr>
mean_bt
<dbl>
lower_bt
<dbl>
upper_bt
<dbl>
letters
<chr>
Soil	5.056803	3.593527	6.986207	b
Glass	2.922635	1.974957	4.172197	a
25% SG	2.275744	1.484350	3.319237	a
75% SG	1.743220	1.080480	2.617077	a
100% SG	2.365977	1.552784	3.438215	a

Treatment_Name
<fctr>
mean_bt
<dbl>
lower_bt
<dbl>
upper_bt
<dbl>
letters
<chr>
Soil	1.5618722	1.1981897	1.985725	c
Glass	0.6096880	0.3811773	0.876005	a
25% SG	1.0336686	0.7449697	1.370132	b
75% SG	0.9237482	0.6506536	1.242025	ab
100% SG	0.8224510	0.5637365	1.123969	ab

Treatment_Name
<fctr>
mean_bt
<dbl>
lower_bt
<dbl>
upper_bt
<dbl>
letters
<chr>
Soil	2.1281031	1.4697164	2.962005	b
Glass	1.9210455	1.3062392	2.699749	b
25% SG	1.0358575	0.6073609	1.578584	a
75% SG	0.8935336	0.4949926	1.398319	a
100% SG	0.8636909	0.4714310	1.360521	a


```{r}
# pick y-positions a bit above the CI max per treatment
pos_agb_R <- letters_agb_R %>%
  mutate(y = pmax(upper_bt, 0) * 1.5)

p1 +
  geom_text(data = pos_agb_R,
            aes(x = Treatment_Name, y = y, label = letters),
            inherit.aes = FALSE, size = 4)

pos_fine_R <- letters_fine_R %>% mutate(y = pmax(upper_bt, 0) * 1.5)
pos_coarse_R <- letters_coarse_R %>% mutate(y = pmax(upper_bt, 0) * 1.5)

fineR<- p2 + geom_text(data = pos_fine_R,
                aes(Treatment_Name, y, label = letters),
                inherit.aes = FALSE, size = 4)

coarseR<- p3 + geom_text(data = pos_coarse_R,
                aes(Treatment_Name, y, label = letters),
                inherit.aes = FALSE, size = 4)
fineR/coarseR
```


LARA
```{r}
fit_agb_L    <- lm(AGB_lp   ~ Treatment_Name, data = biomass_L)
fit_fine_L   <- lm(Fine_lp  ~ Treatment_Name, data = biomass_L)
fit_coarse_L <- lm(Coarse_lp~ Treatment_Name, data = biomass_L)

anova(fit_agb_L); anova(fit_fine_L); anova(fit_coarse_L)

letters_agb_L    <- letters_tbl(fit_agb_L, resp="AGB_lp", transform="exp1")
letters_fine_L   <- letters_tbl(fit_fine_L, resp="Fine_lp", transform="exp1")
letters_coarse_L <- letters_tbl(fit_coarse_L, resp="Coarse_lp", transform="exp1")

```
Analysis of Variance Table

Response: AGB_lp
               Df Sum Sq Mean Sq F value
Treatment_Name  3 2.2464 0.74878  15.677
Residuals      32 1.5284 0.04776        
                  Pr(>F)    
Treatment_Name 1.896e-06 ***
Residuals                   
---
Signif. codes:  
0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
Analysis of Variance Table

Response: Fine_lp
               Df  Sum Sq  Mean Sq F value
Treatment_Name  3 0.39411 0.131370  7.2524
Residuals      32 0.57965 0.018114        
                  Pr(>F)    
Treatment_Name 0.0007607 ***
Residuals                   
---
Signif. codes:  
0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
Analysis of Variance Table

Response: Coarse_lp
               Df  Sum Sq  Mean Sq F value
Treatment_Name  3 0.40375 0.134583  11.717
Residuals      31 0.35608 0.011486        
                 Pr(>F)    
Treatment_Name 2.69e-05 ***
Residuals                  
---
Signif. codes:  
0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

```{r}
p1L <- plot_box_letters(biomass_L, "AGB",       "Above-ground biomass (g)")
p2L <- plot_box_letters(biomass_L, "FineRoot",  "Fine-root biomass (g)")
p3L <- plot_box_letters(biomass_L, "CoarseRoot","Coarse-root biomass (g)")
p1L / p2L / p3L

```

```{r}
letters_agb_L %>% dplyr::select(Treatment_Name, mean_bt, lower_bt, upper_bt, letters) %>% arrange(Treatment_Name)
letters_fine_L %>% dplyr::select(Treatment_Name, mean_bt, lower_bt, upper_bt, letters) %>% arrange(Treatment_Name)
letters_coarse_L %>% dplyr::select(Treatment_Name, mean_bt, lower_bt, upper_bt, letters) %>% arrange(Treatment_Name)

```

Treatment_Name
<fctr>
mean_bt
<dbl>
lower_bt
<dbl>
upper_bt
<dbl>
letters
<chr>
Soil	9.100955e-01	0.5907904597	1.2934917	b
Glass	2.220446e-16	-0.1671670379	0.2007210	a
25% SG	1.924395e-01	-0.0068970857	0.4317871	a
75% SG	2.670008e-01	0.0005011664	0.6044870	a

Treatment_Name
<fctr>
mean_bt
<dbl>
lower_bt
<dbl>
upper_bt
<dbl>
letters
<chr>
Soil	0.30377121	0.164871272	0.4592337	b
Glass	0.00000000	-0.106537049	0.1192406	a
25% SG	0.06074182	-0.052266484	0.1872253	a
75% SG	0.15380027	-0.002367168	0.3344139	ab

Treatment_Name
<fctr>
mean_bt
<dbl>
lower_bt
<dbl>
upper_bt
<dbl>
letters
<chr>
Soil	0.2935711729	0.18237969	0.41521916	b
Glass	0.0003992818	-0.08559203	0.09447725	a
25% SG	0.0473786289	-0.04729395	0.15145904	a
75% SG	0.0244342431	-0.08779547	0.15047173	a
```{r}
# pick y-positions a bit above the CI max per treatment
pos_agb_L <- letters_agb_L %>%
  mutate(y = pmax(upper_bt, 0) * 2.08)

p1L +
  geom_text(data = pos_agb_L,
            aes(x = Treatment_Name, y = y, label = letters),
            inherit.aes = FALSE, size = 4)

pos_fine_L <- letters_fine_L %>% mutate(y = pmax(upper_bt, 0) * 2.08)
pos_coarse_L <- letters_coarse_L %>% mutate(y = pmax(upper_bt, 0) * 2.08)

fine<- p2L + geom_text(data = pos_fine_L,
                aes(Treatment_Name, y, label = letters),
                inherit.aes = FALSE, size = 4)

coarse<- p3L + geom_text(data = pos_coarse_L,
                aes(Treatment_Name, y, label = letters),
                inherit.aes = FALSE, size = 4)
fine/coarse
```



Allometric Coupling: Roots vs Shoots

We evaluate whether substrates shift allocation at a given shoot size (intercept shift) or change scaling (slope).
```{r}
fit_allo_fine   <- lm(Fine_lp   ~ AGB_lp * Treatment_Name + Species, data = biomass)
fit_allo_coarse <- lm(Coarse_lp ~ AGB_lp * Treatment_Name + Species, data = biomass)

anova(fit_allo_fine)
anova(fit_allo_coarse)

# Effects (shifts at mean AGB)
em_fine_shift   <- emmeans(fit_allo_fine,   ~ Treatment_Name | Species, cov.reduce = list(AGB_lp = mean))
em_coarse_shift <- emmeans(fit_allo_coarse, ~ Treatment_Name | Species, cov.reduce = list(AGB_lp = mean))

pairs(em_fine_shift,   adjust = "holm")
pairs(em_coarse_shift, adjust = "holm")

```
Analysis of Variance Table

Response: Fine_lp
                      Df Sum Sq Mean Sq
AGB_lp                 1 8.0734  8.0734
Treatment_Name         4 0.4217  0.1054
Species                1 0.2534  0.2534
AGB_lp:Treatment_Name  4 0.3829  0.0957
Residuals             75 1.0928  0.0146
                       F value    Pr(>F)    
AGB_lp                554.0863 < 2.2e-16 ***
Treatment_Name          7.2358 5.539e-05 ***
Species                17.3933 8.076e-05 ***
AGB_lp:Treatment_Name   6.5692 0.0001376 ***
Residuals                                   
---
Signif. codes:  
0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
Analysis of Variance Table

Response: Coarse_lp
                      Df  Sum Sq Mean Sq
AGB_lp                 1 15.5239 15.5239
Treatment_Name         4  0.7053  0.1763
Species                1  0.1214  0.1214
AGB_lp:Treatment_Name  4  0.1169  0.0292
Residuals             74  1.7454  0.0236
                       F value    Pr(>F)    
AGB_lp                658.1582 < 2.2e-16 ***
Treatment_Name          7.4756 4.099e-05 ***
Species                 5.1490   0.02617 *  
AGB_lp:Treatment_Name   1.2394   0.30169    
Residuals                                   
---
Signif. codes:  
0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
Species = LARA:
 contrast         estimate     SE df t.ratio
 Soil - Glass       0.2182 0.0490 75   4.454
 Soil - 25% SG      0.0258 0.0499 75   0.516
 Soil - 75% SG      0.0147 0.0556 75   0.265
 Soil - 100% SG     0.1898 0.0754 75   2.518
 Glass - 25% SG    -0.1924 0.0400 75  -4.808
 Glass - 75% SG    -0.2034 0.0431 75  -4.718
 Glass - 100% SG   -0.0284 0.0601 75  -0.471
 25% SG - 75% SG   -0.0110 0.0433 75  -0.255
 25% SG - 100% SG   0.1640 0.0599 75   2.737
 75% SG - 100% SG   0.1751 0.0594 75   2.946
 p.value
  0.0002
  1.0000
  1.0000
  0.0696
  0.0001
  0.0001
  1.0000
  1.0000
  0.0464
  0.0300

Species = RHMA:
 contrast         estimate     SE df t.ratio
 Soil - Glass       0.2182 0.0490 75   4.454
 Soil - 25% SG      0.0258 0.0499 75   0.516
 Soil - 75% SG      0.0147 0.0556 75   0.265
 Soil - 100% SG     0.1898 0.0754 75   2.518
 Glass - 25% SG    -0.1924 0.0400 75  -4.808
 Glass - 75% SG    -0.2034 0.0431 75  -4.718
 Glass - 100% SG   -0.0284 0.0601 75  -0.471
 25% SG - 75% SG   -0.0110 0.0433 75  -0.255
 25% SG - 100% SG   0.1640 0.0599 75   2.737
 75% SG - 100% SG   0.1751 0.0594 75   2.946
 p.value
  0.0002
  1.0000
  1.0000
  0.0696
  0.0001
  0.0001
  1.0000
  1.0000
  0.0464
  0.0300

P value adjustment: holm method for 10 tests 
Species = LARA:
 contrast         estimate     SE df t.ratio
 Soil - Glass     -0.19936 0.0622 74  -3.204
 Soil - 25% SG    -0.02410 0.0636 74  -0.379
 Soil - 75% SG    -0.00421 0.0708 74  -0.060
 Soil - 100% SG    0.07098 0.0950 74   0.747
 Glass - 25% SG    0.17527 0.0514 74   3.412
 Glass - 75% SG    0.19515 0.0552 74   3.537
 Glass - 100% SG   0.27034 0.0758 74   3.565
 25% SG - 75% SG   0.01988 0.0556 74   0.358
 25% SG - 100% SG  0.09508 0.0758 74   1.255
 75% SG - 100% SG  0.07519 0.0751 74   1.002
 p.value
  0.0140
  1.0000
  1.0000
  1.0000
  0.0084
  0.0064
  0.0064
  1.0000
  1.0000
  1.0000

Species = RHMA:
 contrast         estimate     SE df t.ratio
 Soil - Glass     -0.19936 0.0622 74  -3.204
 Soil - 25% SG    -0.02410 0.0636 74  -0.379
 Soil - 75% SG    -0.00421 0.0708 74  -0.060
 Soil - 100% SG    0.07098 0.0950 74   0.747
 Glass - 25% SG    0.17527 0.0514 74   3.412
 Glass - 75% SG    0.19515 0.0552 74   3.537
 Glass - 100% SG   0.27034 0.0758 74   3.565
 25% SG - 75% SG   0.01988 0.0556 74   0.358
 25% SG - 100% SG  0.09508 0.0758 74   1.255
 75% SG - 100% SG  0.07519 0.0751 74   1.002
 p.value
  0.0140
  1.0000
  1.0000
  1.0000
  0.0084
  0.0064
  0.0064
  1.0000
  1.0000
  1.0000

P value adjustment: holm method for 10 tests 


Allometry plots (log scale)
```{r}
#this is the old version of the code for plots with the legend next to each one
# plot_allometry <- function(df, y, ylab){
#     ggplot(df, aes(AGB_lp, .data[[y]], color = Treatment_Name)) +
#         geom_point(alpha = 0.6) +
#         geom_smooth(method = "lm", se = FALSE) +
#         scale_color_manual(values = color_vals) +
#         labs(x = "log1p(AGB)", y = ylab, color = "Treatment") +
#         theme(legend.position = "right")
# }

plot_allometry <- function(df, y, ylab){
  ggplot(df, aes(AGB_lp, .data[[y]], color = Treatment_Name)) +
    geom_point(alpha = 0.6) +
    geom_smooth(method = "lm", se = FALSE) +
    scale_color_manual(
      values = color_vals,
      name   = "Treatment",
      breaks = names(color_vals)   # ensure identical order across plots
    ) +
    guides(color = guide_legend(nrow = 1, byrow = TRUE)) +  # one-row legend
    labs(x = "log1p(AGB)", y = ylab) +
    theme_classic() +
    theme(legend.position = "right")
}


# Subset data by species
rhma_data <- biomass %>% filter(Species == "RHMA")
lara_data <- biomass %>% filter(Species == "LARA")

# ---- R. mangle ----
pA_RHMA <- plot_allometry(rhma_data, "Fine_lp",   "log1p(Fine root)")
pB_RHMA <- plot_allometry(rhma_data, "Coarse_lp", "log1p(Coarse root)")

# ---- L. racemosa ----
pA_LARA <- plot_allometry(lara_data, "Fine_lp",   "log1p(Fine root)")
pB_LARA <- plot_allometry(lara_data, "Coarse_lp", "log1p(Coarse root)")

# # Optional: combine plots
# library(patchwork)
# (pA_RHMA | pB_RHMA) / (pA_LARA | pB_LARA) +
#   plot_annotation(tag_levels = "A") &
#   theme(plot.tag = element_text(face = "bold", size = 14))

library(patchwork)

combined_plot <-
  (pA_LARA | pB_LARA) /
  (pA_RHMA | pB_RHMA) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(face = "bold", size = 14))

combined_plot2<- combined_plot +
  plot_layout(guides = "collect") &
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.box = "horizontal",
    legend.title = element_text(face = "bold"),
    legend.margin = margin(t = 2, r = 0, b = 2, l = 0)
  )
combined_plot2


pA <- plot_allometry(biomass, "Fine_lp",   "log1p(Fine root)")
pB <- plot_allometry(biomass, "Coarse_lp", "log1p(Coarse root)")
pA + pB + plot_layout(guides = "collect")



```

Allocation as Log-Ratios (Root:Shoot and Fine:Coarse)

Ratios are analyzed as differences of logs to stabilize variance; results are back-transformed for presentation.
```{r}
fit_logRS <- lm(logRS ~ Treatment_Name * Species, data = biomass)
fit_logFC <- lm(logFC ~ Treatment_Name * Species, data = biomass)

anova(fit_logRS); anova(fit_logFC)

emRS <- emmeans(fit_logRS, ~ Treatment_Name | Species)
emFC <- emmeans(fit_logFC, ~ Treatment_Name | Species)

pairs(emRS, adjust = "holm")
pairs(emFC, adjust = "holm")

# Letters (per species), back-transform to ratio scale
letters_RS_R <- letters_tbl(lm(logRS ~ Treatment_Name, data = biomass_R), resp="logRS", transform="exp1")
letters_RS_L <- letters_tbl(lm(logRS ~ Treatment_Name, data = biomass_L), resp="logRS", transform="exp1")
letters_FC_R <- letters_tbl(lm(logFC ~ Treatment_Name, data = biomass_R), resp="logFC", transform="exp1")
letters_FC_L <- letters_tbl(lm(logFC ~ Treatment_Name, data = biomass_L), resp="logFC", transform="exp1")

```
Analysis of Variance Table

Response: logRS
                       Df  Sum Sq  Mean Sq
Treatment_Name          4 0.65888 0.164719
Species                 1 0.00303 0.003028
Treatment_Name:Species  3 0.10855 0.036183
Residuals              76 1.86726 0.024569
                       F value    Pr(>F)    
Treatment_Name          6.7043 0.0001123 ***
Species                 0.1232 0.7265147    
Treatment_Name:Species  1.4727 0.2286868    
Residuals                                   
---
Signif. codes:  
0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
Analysis of Variance Table

Response: logFC
                       Df Sum Sq Mean Sq
Treatment_Name          4 1.4276 0.35689
Species                 1 1.0944 1.09440
Treatment_Name:Species  3 0.9359 0.31196
Residuals              76 3.3897 0.04460
                       F value    Pr(>F)    
Treatment_Name          8.0017 1.942e-05 ***
Species                24.5372 4.296e-06 ***
Treatment_Name:Species  6.9945 0.0003233 ***
Residuals                                   
---
Signif. codes:  
0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
Species = LARA:
 contrast         estimate     SE df t.ratio
 Soil - Glass     -0.19193 0.0701 76  -2.738
 Soil - 25% SG    -0.10018 0.0720 76  -1.391
 Soil - 75% SG    -0.11375 0.0809 76  -1.405
 Soil - 100% SG     nonEst     NA NA      NA
 Glass - 25% SG    0.09175 0.0720 76   1.274
 Glass - 75% SG    0.07818 0.0809 76   0.966
 Glass - 100% SG    nonEst     NA NA      NA
 25% SG - 75% SG  -0.01357 0.0826 76  -0.164
 25% SG - 100% SG   nonEst     NA NA      NA
 75% SG - 100% SG   nonEst     NA NA      NA
 p.value
  0.0462
  0.8200
  0.8200
      NA
  0.8200
  0.8200
      NA
  0.8700
      NA
      NA

Species = RHMA:
 contrast         estimate     SE df t.ratio
 Soil - Glass     -0.14871 0.0701 76  -2.121
 Soil - 25% SG    -0.18596 0.0701 76  -2.653
 Soil - 75% SG    -0.27616 0.0701 76  -3.940
 Soil - 100% SG   -0.00842 0.0701 76  -0.120
 Glass - 25% SG   -0.03725 0.0701 76  -0.531
 Glass - 75% SG   -0.12744 0.0701 76  -1.818
 Glass - 100% SG   0.14029 0.0701 76   2.001
 25% SG - 75% SG  -0.09020 0.0701 76  -1.287
 25% SG - 100% SG  0.17753 0.0701 76   2.533
 75% SG - 100% SG  0.26773 0.0701 76   3.819
 p.value
  0.2229
  0.0777
  0.0018
  1.0000
  1.0000
  0.2920
  0.2447
  0.6063
  0.0937
  0.0024

P value adjustment: holm method for varying numbers of tests 
Species = LARA:
 contrast         estimate     SE df t.ratio
 Soil - Glass      0.00825 0.0944 76   0.087
 Soil - 25% SG    -0.01138 0.0970 76  -0.117
 Soil - 75% SG    -0.11107 0.1090 76  -1.018
 Soil - 100% SG     nonEst     NA NA      NA
 Glass - 25% SG   -0.01963 0.0970 76  -0.202
 Glass - 75% SG   -0.11932 0.1090 76  -1.094
 Glass - 100% SG    nonEst     NA NA      NA
 25% SG - 75% SG  -0.09969 0.1110 76  -0.896
 25% SG - 100% SG   nonEst     NA NA      NA
 75% SG - 100% SG   nonEst     NA NA      NA
 p.value
  1.0000
  1.0000
  1.0000
      NA
  1.0000
  1.0000
      NA
  1.0000
      NA
      NA

Species = RHMA:
 contrast         estimate     SE df t.ratio
 Soil - Glass      0.39621 0.0944 76   4.195
 Soil - 25% SG    -0.19861 0.0944 76  -2.103
 Soil - 75% SG    -0.21552 0.0944 76  -2.282
 Soil - 100% SG   -0.17731 0.0944 76  -1.877
 Glass - 25% SG   -0.59483 0.0944 76  -6.298
 Glass - 75% SG   -0.61173 0.0944 76  -6.477
 Glass - 100% SG  -0.57352 0.0944 76  -6.072
 25% SG - 75% SG  -0.01691 0.0944 76  -0.179
 25% SG - 100% SG  0.02130 0.0944 76   0.226
 75% SG - 100% SG  0.03821 0.0944 76   0.405
 p.value
  0.0005
  0.1939
  0.1518
  0.2572
  <.0001
  <.0001
  <.0001
  1.0000
  1.0000
  1.0000

P value adjustment: holm method for varying numbers of tests 


Figures: Back-transformed ratios with letters
```{r}
plot_ratio_letters <- function(letters_df, title, ylab){
  letters_df %>%
    ggplot(aes(Treatment_Name, mean_bt)) +
    geom_point(size = 3) +
    geom_errorbar(aes(ymin = lower_bt, ymax = upper_bt), width = 0.15) +
    geom_text(aes(label = letters, y = upper_bt*1.05), vjust = 0.25, size = 4) +
    labs(x = NULL, y = ylab, title = title) +
    theme(axis.text.x = element_text(angle = 15, hjust = 1))
}

(p_RS_R <- plot_ratio_letters(letters_RS_R, "RHMA — Root:Shoot (back-transformed)", "Root:Shoot ratio"))
(p_RS_L <- plot_ratio_letters(letters_RS_L, "LARA — Root:Shoot (back-transformed)", "Root:Shoot ratio"))
(p_FC_R <- plot_ratio_letters(letters_FC_R, "RHMA — Fine:Coarse (back-transformed)", "Fine:Coarse ratio"))
(p_FC_L <- plot_ratio_letters(letters_FC_L, "LARA — Fine:Coarse (back-transformed)", "Fine:Coarse ratio"))

p_RS_L/p_RS_R
p_FC_L/p_FC_R

```

Model Diagnostics (brief)
```{r}
check_model <- function(fit){
  list(
    check_model = performance::check_model(fit),
    r2          = performance::r2(fit)
  )
}

diag_list <- list(
  agb_R    = check_model(fit_agb_R),
  fine_R   = check_model(fit_fine_R),
  coarse_R = check_model(fit_coarse_R),
  agb_L    = check_model(fit_agb_L),
  fine_L   = check_model(fit_fine_L),
  coarse_L = check_model(fit_coarse_L),
  allo_fine   = check_model(fit_allo_fine),
  allo_coarse = check_model(fit_allo_coarse),
  logRS = check_model(fit_logRS),
  logFC = check_model(fit_logFC)
)
diag_list$r2

```

Tables for Manuscript
6.1 Estimated means (back-transformed) with Holm letters
```{r}
fmt_letters <- function(df, digits = 2){
  df %>%
    transmute(
      Treatment = Treatment_Name,
      Mean = round(mean_bt, digits),
      `95% CI` = sprintf("[%s, %s]", round(lower_bt, digits), round(upper_bt, digits)),
      Letters = letters
    )
}

list(
  RHMA_AGB    = fmt_letters(letters_agb_R),
  RHMA_Fine   = fmt_letters(letters_fine_R),
  RHMA_Coarse = fmt_letters(letters_coarse_R),
  LARA_AGB    = fmt_letters(letters_agb_L),
  LARA_Fine   = fmt_letters(letters_fine_L),
  LARA_Coarse = fmt_letters(letters_coarse_L),
  RHMA_RootShoot = fmt_letters(letters_RS_R),
  LARA_RootShoot = fmt_letters(letters_RS_L),
  RHMA_FineCoarse = fmt_letters(letters_FC_R),
  LARA_FineCoarse = fmt_letters(letters_FC_L)
)

```

Pairwise contrasts (Holm)
```{r}
contrasts_agb_R    <- pairs(emmeans(fit_agb_R,    ~ Treatment_Name), adjust="holm") %>% broom::tidy()
contrasts_fine_R   <- pairs(emmeans(fit_fine_R,   ~ Treatment_Name), adjust="holm") %>% broom::tidy()
contrasts_coarse_R <- pairs(emmeans(fit_coarse_R, ~ Treatment_Name), adjust="holm") %>% broom::tidy()

contrasts_agb_L    <- pairs(emmeans(fit_agb_L,    ~ Treatment_Name), adjust="holm") %>% broom::tidy()
contrasts_fine_L   <- pairs(emmeans(fit_fine_L,   ~ Treatment_Name), adjust="holm") %>% broom::tidy()
contrasts_coarse_L <- pairs(emmeans(fit_coarse_L, ~ Treatment_Name), adjust="holm") %>% broom::tidy()

contrasts_RS <- pairs(emRS, adjust="holm") %>% broom::tidy()
contrasts_FC <- pairs(emFC, adjust="holm") %>% broom::tidy()

list(
  RHMA_AGB    = contrasts_agb_R,
  RHMA_Fine   = contrasts_fine_R,
  RHMA_Coarse = contrasts_coarse_R,
  LARA_AGB    = contrasts_agb_L,
  LARA_Fine   = contrasts_fine_L,
  LARA_Coarse = contrasts_coarse_L,
  RootShoot   = contrasts_RS,
  FineCoarse  = contrasts_FC
)

```

Figure Caption Templates (edit in manuscript)

Figure X. RHMA biomass by substrate. Above-ground (AGB), fine-root, and coarse-root biomass for Rhizophora mangle. Points show individuals, boxes show distribution; black diamonds mark means. Letters denote Holm-adjusted groups from linear mixed-effects models on log1p-transformed responses with a random Site effect; values are back-transformed for interpretation.

Figure Y. LARA biomass by substrate. As in Fig. X for Laguncularia racemosa.

Figure Z. Root–shoot allometry by substrate. log1p(root component) vs log1p(AGB) with per-treatment regression lines; vertical shifts indicate allocation differences at a common shoot size, and slope differences indicate size-dependent scaling. Tests from mixed-effects models including Species and a random Site intercept.

Figure W. Allocation ratios. Back-transformed root:shoot and fine:coarse mean ratios (±95% CI) by substrate, with Holm letters from mixed-effects models on log-ratios.



