---
title: "Mangrove Predation Analysis"
author: "Sarai Hutchinson"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cerulean
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)

# Load required libraries
library(tidyverse)
library(ggplot2)
library(dplyr)
library(stringr)
library(purrr)
library(scales)

# Define Okabe-Ito colorblind-friendly palette
okabe_ito <- c("#009E73", "#D55E00", "#E69F00", "#56B4E9", "#F0E442", "#0072B2", "#CC79A7", "#999999")

# Define shapes for each Treatment_Name
shape_vals <- c(
  "Soil"                = 0,  # open square
  "Crushed glass"       = 17, # filled triangle
  "25% SG"              = 18, # filled diamond
  "75% SG"              = 8,  # asterisk
  "100% Sargassum (SG)" = 15  # filled square
)

# Define colors using okabe_ito palette
color_vals <- c(
  "Soil"                = okabe_ito[1], # green  (#009E73)
  "Crushed glass"       = okabe_ito[6], # blue   (#0072B2)
  "25% SG"              = okabe_ito[3], # orange (#E69F00)
  "75% SG"              = okabe_ito[2], # vermillion (#D55E00)
  "100% Sargassum (SG)" = okabe_ito[7]  # purple (#CC79A7)
)
```

# Introduction

This analysis examines the predation patterns of red (*Rhizophora mangle*, RHMA) and white (*Laguncularia racemosa*, LARA) mangrove seedlings under different substrate treatments.

## Treatments

1. **Soil**: Traditional potting soil (control)
2. **Crushed glass**: 100% crushed glass substrate
3. **25% SG**: 3:1 ratio of crushed glass to *Sargassum*
4. **75% SG**: 1:3 ratio of crushed glass to *Sargassum*
5. **100% Sargassum (SG)**: Pure *Sargassum* substrate

```{r data-import}
# Read data
ALLDATA <- read_csv("C:\\Users\\sarai\\OneDrive\\Email attachments\\Documents\\UVI\\Thesis.Work\\SargMangs\\Data/Hutchinson_MonitoringAndIntake_Sites.csv")

# Clean treatment labels
ALLDATA <- ALLDATA %>%
  mutate(
    Treatment_Name = case_when(
      Treatment == "A" ~ "Soil",
      Treatment == "B" ~ "Crushed glass",
      Treatment == "C" ~ "100% Sargassum (SG)",
      Treatment == "D" ~ "25% SG",
      Treatment == "E" ~ "75% SG",
      TRUE ~ as.character(Treatment)
    ),
    Treatment_Name = factor(
      Treatment_Name,
      levels = c("Soil", "Crushed glass", "25% SG", "75% SG", "100% Sargassum (SG)")
    )
  )

# Derive Sargassum indicators
ALLDATA <- ALLDATA %>%
  mutate(
    Sarg_present = if_else(
      Treatment_Name %in% c("75% SG", "25% SG", "100% Sargassum (SG)"),
      1L, 0L
    ),
    Sarg_group = factor(if_else(Sarg_present == 1L, "With Sargassum", "No Sargassum"),
                        levels = c("No Sargassum","With Sargassum")),
    SG_pct = case_when(
      Treatment_Name == "Soil" ~ 0,
      Treatment_Name == "Crushed glass" ~ 0,
      Treatment_Name == "25% SG" ~ 25,
      Treatment_Name == "75% SG" ~ 75,
      Treatment_Name == "100% Sargassum (SG)"~ 100,
      TRUE ~ NA_real_
    )
  )

# Create binary survival variable and factorize
ALLDATA_SurvBin <- ALLDATA %>%
  mutate(
    Survival = ifelse(Survivorship %in% c("A", "DY"), 1, 0),
    Time_Period = factor(Time_Period, levels = c("0", "1", "2", "3", "4", "5", "6")),
    Species = factor(Species)
  )

# Quality control - remove resurrection cases
qc_resurrections <- ALLDATA_SurvBin %>%
  mutate(TP_num = as.integer(as.character(Time_Period))) %>%
  arrange(Species, Prop_ID, TP_num) %>%
  group_by(Species, Prop_ID) %>%
  mutate(change = Survival - lag(Survival)) %>%
  filter(change == 1 & lag(Survival) == 0) %>%
  ungroup()

flagged_ids <- qc_resurrections %>% distinct(Species, Prop_ID)

ALLDATA_clean <- ALLDATA_SurvBin %>%
  anti_join(flagged_ids, by = c("Species","Prop_ID"))

cat("Data cleaned. Removed", nrow(flagged_ids), "plants with resurrection issues.\n")

```
I want to look at predation levels across treatments overtime by looking at the Condition and Pests columns.
```{r}
# define your lookup once
cond_map <- c(H = "Healthy", P = "Predation", UH = "Unhealthy")

ALLDATA_pred <- ALLDATA_clean %>%
  mutate(
    condition_raw = as.character(Condition),

    # normalize: uppercase + unify separators
    .norm = condition_raw %>%
      str_to_upper() %>%
      str_replace_all("[,;/\\|&\\+]+", " ") %>%
      str_squish(),

    # split to tokens, keep only valid codes
    .tokens = str_split(.norm, "\\s+"),
    .tokens = map(.tokens, ~ .x[.x %in% names(cond_map)]),

    # flags
    has_healthy   = map_lgl(.tokens, ~ "H"  %in% .x),
    has_predation = map_lgl(.tokens, ~ "P"  %in% .x),
    has_unhealthy = map_lgl(.tokens, ~ "UH" %in% .x),

    pred_any = as.integer(has_predation),

    overall_health = case_when(
      has_unhealthy                ~ "Unhealthy",
      has_healthy & !has_unhealthy ~ "Healthy",
      TRUE                         ~ NA_character_
    ),

    # map codes -> labels (no recode call)
    .mapped = map(.tokens, ~ unname(cond_map[.x])),

    # order for readability
    .mapped = map(.mapped, ~ {
      ord <- c("Predation","Unhealthy","Healthy")
      unique(.x[order(factor(.x, levels = ord))])
    }),

    Condition_std = map_chr(.mapped, ~ if (length(.x)) paste(.x, collapse = ", ") else NA_character_)
  ) %>%
  dplyr::select(-.norm, -.tokens, -.mapped)
```

```{r}
pred_summary <- ALLDATA_pred %>%
  group_by(Species, Treatment_Name, Time_Period) %>%
  summarise(
    n = n(),
    n_pred = sum(pred_any == 1, na.rm = TRUE),
    pct_pred = 100 * n_pred / n,
    .groups = "drop"
  )

```

```{r}
library(lme4)

pred_glmm <- ALLDATA_pred %>%
  filter(!is.na(pred_any)) %>%
  mutate(
    Species        = factor(Species),
    Treatment_Name = factor(Treatment_Name),
    Time_Period    = as.numeric(Time_Period) # or factor(Time_Period)
  )

#Make separate glmer's by Species#
fit <- glmer(pred_any ~ Species * Treatment_Name + Time_Period +
               (1|Site) + (1|Prop_ID),
             data = pred_glmm, family = binomial,
             control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5)))
print(fit)

library(car)
car::Anova(fit, type = "III")  # test fixed terms

library(emmeans)
emm_trt_by_sp <- emmeans(fit, ~ Treatment_Name | Species, type = "response")
pairs(emm_trt_by_sp, adjust = "bonferroni")  # treatment differences within species
```

```{r}
# starting from your fitted model `fit`
emm_trt_by_sp <- emmeans(fit, ~ Treatment_Name | Species, type = "response")

# Pairwise treatment contrasts within each species, Holm-adjusted
pairs_holm <- contrast(emm_trt_by_sp, method = "pairwise", adjust = "holm") %>%
  summary(infer = TRUE)    # adds CIs; tests are on log-odds scale

# Split out tables for quick reading
pairs_holm_rhma <- as.data.frame(pairs_holm) %>% filter(Species == "RHMA")
pairs_holm_lara <- as.data.frame(pairs_holm) %>% filter(Species == "LARA")

# Check if RHMA has any significant differences (Holm-adjusted)
any_sig_rhma <- any(pairs_holm_rhma$p.value < 0.05, na.rm = TRUE)
any_sig_rhma
pairs_holm_rhma   # inspect

# (Optional) compact letters display per species, Holm-adjusted
cld_lara <- cld(emm_trt_by_sp, by = "Species", adjust = "holm") %>%
  as.data.frame() %>% filter(Species == "LARA")
cld_rhma <- cld(emm_trt_by_sp, by = "Species", adjust = "holm") %>%
  as.data.frame() %>% filter(Species == "RHMA")

cld_lara
cld_rhma

pairs_holm <- contrast(emm_trt_by_sp, method = "pairwise", adjust = "holm") %>% 
  as.data.frame()

sig_pairs <- pairs_holm %>% 
  filter(p.value < 0.05) %>% 
  arrange(Species, p.value)

sig_pairs
```

```{r}
# 1) EMMeans on probability scale
emm_trt_by_sp <- emmeans(fit, ~ Treatment_Name | Species, type = "response")

emm_df <- as.data.frame(emm_trt_by_sp) %>%
  dplyr::rename(
    prob = prob,              # <- was "response"
    lwr  = asymp.LCL,
    upr  = asymp.UCL
  ) %>%
  dplyr::mutate(Treatment_Name = factor(Treatment_Name, levels = names(color_vals))) %>%
  dplyr::group_by(Species) %>%
  dplyr::arrange(prob, .by_group = TRUE) %>%
  dplyr::ungroup() %>%
  # lock legend/order to your palette keys
  mutate(Treatment_Name = factor(Treatment_Name, levels = names(color_vals))) %>%
  # order within each species by estimated prob (for nicer stacking in coord_flip)
  group_by(Species) %>%
  arrange(prob, .by_group = TRUE) %>%
  ungroup()

# 2) Plot
ggplot(emm_df,
       aes(x = Treatment_Name, y = prob,
           color = Treatment_Name, shape = Treatment_Name)) +
  geom_errorbar(aes(ymin = lwr, ymax = upr), width = 0.2, alpha = 0.8) +
  geom_point(size = 2.6) +
  facet_wrap(~ Species, nrow = 1) +
  coord_flip() +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  scale_color_manual(values = color_vals, name = "Treatment") +
  scale_shape_manual(values = shape_vals,  name = "Treatment") +
  labs(
    title = "Predation probability by treatment (EMMeans ±95% CI)",
    x = NULL, y = "Predation probability"
  ) +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",
    panel.grid.minor = element_blank()
  )
```


